---
title: "Bayesian Multilevel Models in R with `MCMCglmm`"
output:
  html_notebook:
    toc: yes
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">#")
```

In this note I'll replicate the models I demonstrated in the class with
Bayesian estimation and the `MCMCglmm` package. 

# 1. Importing Data Into R

We'll import the data in .sav format. 

```{r import_sav, message=FALSE}
# install.packages("tidyverse")  # if you haven't installed the `tidyverse` package
library(tidyverse)
theme_set(theme_bw())  # Use a different theme for plotting with ggplot2
library(haven)
# Read in the data (pay attention to the directory)
hsball <- read_sav("../data/hsball.sav")
hsball  # print the first 10 rows
```

# 2. Random Intercept Model

First load the package:

```{r load_lme4, message=FALSE}
# install.packages("MCMCglmm")  # if you haven't installed the `MCMCglmm`
# package, which requires `rstan` to be installed first
library(MCMCglmm)
```

```{r ran_int, results='hide'}
ran_int <- MCMCglmm(fixed = mathach ~ 1, 
                    random = ~ idv(1):id,  # it's the way MCMCglmm specify a random intercept
                    data = hsball, 
                    prior = list(B = list(mu = 0, V = 1000),  # N(0, 1000) prior for intercept
                                 G = list(G1 = list(V = 1, nu = 1)),  # weakly informative prior for tau00
                                 R = list(V = 1, nu = 0.002)))  # vague prior for sigma-sq
```

You can summarize the results using

```{r summary_ran_int}
summary(ran_int)
```

## Plot Posterior Density

```{r}
plot(ran_int)
```


# 3. Model With a Lv-2 predictor

```{r m_lv2, results='hide'}
m_lv2 <- MCMCglmm(fixed = mathach ~ meanses, 
                  random = ~ idv(1):id,  # it's the way MCMCglmm specify a random intercept
                  data = hsball, 
                  prior = list(# N(0, 100) prior for coefficient
                    B = list(mu = c(0, 0), V = diag(c(1000, 100))),   
                    G = list(G1 = list(V = 1, nu = 1)),  # weakly informative prior for tau00
                    R = list(V = 1, nu = 0.002)  # vague prior for sigma-sq
                  ))
```

```{r summary_m_lv2}
summary(m_lv2)
```

# 4. Decomposing Effects

```{r contextual, results='hide'}
hsball <- hsball %>% 
  group_by(id) %>%   # operate within schools
  mutate(ses_cm = mean(ses),   # create cluster means (the same as `meanses`)
         ses_cmc = ses - ses_cm) %>%   # cluster-mean centered
  ungroup()  # exit the "editing within groups" mode
decompose <- MCMCglmm(fixed = mathach ~ meanses + ses_cmc, 
                      random = ~ idv(1):id,  # it's the way MCMCglmm specify a random intercept
                      data = hsball, 
                      prior = list(# N(0, 100) prior for coefficient
                        B = list(mu = c(0, 0, 0), V = diag(c(1000, 100, 100))),   
                        G = list(G1 = list(V = 1, nu = 1)),  # weakly informative prior for tau00
                        R = list(V = 1, nu = 0.002)  # vague prior for sigma-sq
                      ))
```

```{r summary_contextual}
summary(decompose)
```

# 5. Run the Random-Coefficients Model

```{r ran_slp, results='hide'}
ran_slp <- MCMCglmm(fixed = mathach ~ meanses + ses_cmc, 
                    # Unstructured level-2 covariance matrix
                    random = ~ us(1 + ses_cmc):id,  
                    data = hsball, 
                    prior = list(# N(0, 100) prior for coefficient
                      B = list(mu = c(0, 0, 0), V = diag(c(1000, 100, 100))), 
                      # weakly informative prior for tau00, tau01, and tau11
                      G = list(G1 = list(V = diag(2), nu = 2)),
                      R = list(V = 1, nu = 0.002)  # vague prior for sigma-sq
                    ))
```

```{r summary_ran_slp}
summary(ran_slp)
```

# 6. Analyzing Cross-Level Interactions

```{r crlv_int, results='hide'}
crlv_int <- MCMCglmm(fixed = mathach ~ meanses + sector * ses_cmc, 
                     # Unstructured level-2 covariance matrix
                     random = ~ us(1 + ses_cmc):id,  
                     data = hsball, 
                     prior = list(# N(0, 25) prior for interaction
                      B = list(mu = c(0, 0, 0, 0, 0), 
                               V = diag(c(1000, 100, 100, 100, 25))), 
                      # weakly informative prior for tau00, tau01, and tau11
                      G = list(G1 = list(V = diag(2), nu = 2)),
                      R = list(V = 1, nu = 0.002)  # vague prior for sigma-sq
                    ))
```

```{r summary_crlv_int}
summary(crlv_int)
```
