---
title: "Computing Effect Size for Multilevel Models in R"
date: "`r format(Sys.time(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = ">#")
```

## Importing Data Into R

We'll import the data in .sav format. 

```{r import_sav, message=FALSE}
# install.packages("tidyverse")  # if you haven't installed the `tidyverse` package
library(tidyverse)
theme_set(theme_bw())  # Use a different theme for plotting with ggplot2
library(haven)
# Read in the data (pay attention to the directory)
hsball <- read_sav("../data/hsball.sav")
```

## Standardized Mean Difference (SMD)

### Fit a model with `sector` as predictor

```{r load_lme4, message=FALSE}
library(lme4)
```

```{r m_sector}
m_sector <- lmer(mathach ~ sector + (1 | id), data = hsball)
```

### Obtain $d$ statistic (and 95% CI)

To compute SMD for a binary predictor in a two-level model, using the approach
by Hedges (2009), please use the provided R script file `mlm_d.R`.

```{r mlm_d.R}
source("mlm_d.R")
mlm_d
```

After you source the R script, simply call the function `mlm_d.R` on the model:

```{r}
mlm_d(m_sector)
```

## Variance Accounted For ($R^2$)

### Fit a cross-level interaction model (just an example)

```{r crlv_int}
hsball <- hsball %>% 
  group_by(id) %>%   # operate within schools
  mutate(ses_cm = mean(ses),   # create cluster means (the same as `meanses`)
         ses_cmc = ses - ses_cm) %>%   # cluster-mean centered
  ungroup()  # exit the "editing within groups" mode
crlv_int <- lmer(mathach ~ meanses + sector * ses_cmc + (ses_cmc | id), 
                 data = hsball)
```

### Obtain $R^2$:

Use the `sjstats::r2()` function

```{r}
sjstats::r2(crlv_int)
library(sjstats)
r2(crlv_int)
```

Generally you should report the `Marginal R2` value. Therefore, the model 
accounted for `r sjstats::r2(crlv_int)$rsq.marginal`% of the variance of 
math achievement scores

### Obtain 95% CI

This can be done with parametric bootstrap (`lme4::bootMer()`), but will take
a few minutes:

```{r, warning = FALSE, message=FALSE}
mlm_r2 <- function(x, ci = TRUE, 
                   nsim = 999,  # Some replications may fail, so double check
                   .progress = "txt", ...) {
  boot_r2 <- lme4::bootMer(x, function(x) sjstats::r2(x)$rsq.marginal, 
                           nsim = nsim, .progress = .progress, ...)
  cat("\nCompleted", length(boot_r2$t), "bootstrap samples\n")
  boot_r2$R <- length(boot_r2$t)
  boot_r2_ci <- boot::boot.ci(boot_r2, type = "perc")$perc[4:5]
  setNames(c(boot_r2$t0, boot_r2_ci[1], boot_r2_ci[2]), 
           c("Marginal R2", "ci_ll", "ci_ul"))
}
set.seed(123)
mlm_r2(crlv_int)
```

